Advancements in Moltres for Time-Dependent Multiphysics Molten Salt Reactor Modeling

### 2025-01-03 introduction.tex
Edits before format review

The \gls{MSR} is one of six advanced reactor concepts selected for improved safety, sustainability,
efficiency, and cost over the current generation of predominantly \glspl{LWR} at the \gls{GIF}
\cite{u.s._doe_nuclear_energy_research_advisory_committee_technology_2002}.
Due to the high thermal expansion coefficient, \glspl{MSR} possess an inherently robust
safety feature in the strong negative fuel temperature coefficient of
reactivity \cite{elsheikh_safety_2013}. This reactivity coefficient limits the
maximum temperature that the reactor core would experience in an accident
scenario such as an unprotected reactivity insertion because the subsequent
rise in core temperatures induces a significant drop in reactivity,
quickly neutralizing the initial reactivity insertion. \glspl{MSR} also
operate at a large thermal margin to boiling and can rely on natural
circulation in the event of a pump failure. For an additional layer of safety, many \gls{MSR}
designs incorporate a drain plug made of actively-cooled frozen salt, which
melts when the core temperatures exceed safety thresholds. Hot molten salt
in the core would then flow into a drain tank designed to hold the fuel salt in
a subcritical configuration to disrupt any further chain fission reactions.

@@

\gls{ORNL} researchers first conceived the \gls{MSR} concept in pursuit of a high-temperature
liquid fuel reactor for the US Aircraft Nuclear Propulsion program in
the 1950s \cite{rosenthal_molten-salt_1970}. They
built the first ever operational \gls{MSR}, the 2.5 MW$_{\text{th}}$
\gls{ARE} reactor at \gls{ORNL}. The successful demonstration of the \gls{ARE} spurred further
research into adapting \glspl{MSR} for civilian power generation \cite{rosenthal_molten-salt_1970}.
Continued \gls{MSR} research efforts culminated in designing constructing, and successfully
operating of the 8-MW$_{\text{th}}$, thermal-spectrum \gls{MSRE} with
graphite channels and a LiF-BeF$_2$-ZrF$_4$-UF$_4$ fuel salt mixture
\cite{haubenreich_experience_1970}. In addition to other operational achievements, the
\gls{MSRE} became the first reactor to run on $^{233}$U fuel bred from $^{232}$Th. Building on
their experience with the \gls{MSRE}, \gls{ORNL} proposed a new program for constructing and
operating a demonstration reactor based on the \gls{MSBR} concept, which they had also developed
\cite{macpherson_molten_1985}. The \gls{MSBR} is a thermal-spectrum \gls{MSR} with fertile
$^{232}$Th isotopes mixed directly into the \gls{FLiBe} molten salt for $^{233}$U breeding. The
\gls{MSBR} was to rely on continuous online reprocessing to add fertile
material and remove fission product neutron poisons.
However, the \gls{MSBR} project was canceled prior to the demonstration stage in
favor of the \gls{LMFBR}, which had benefited from a head start in development and more substantial
political backing \cite{macpherson_molten_1985}.

Following a relative lull lasting until the late 1990s, renewed research efforts and interest from
the \gls{GIF}
provided new impetus for \gls{MSR} research and development. As of the end of 2022, numerous
\gls{MSR} designs exist at various stages of development. Leading \gls{MSR} designs, in terms of
development, licensing, or demonstration, include the \gls{MSFR} \cite{merle_optimized_2007},
\gls{MCFR} \cite{terrapower_terrapower_2021}, TMSR-LF1 \cite{zhang_review_2018}, and \gls{IMSR}
\cite{leblanc_18_2017}. The \gls{MSFR} is a fast-spectrum breeder reactor developed through
collaboration among European institutes with funding support from the
European Union. Figure \ref{fig:msfr} shows a schematic diagram of the \gls{MSFR}. As opposed to
the multi-channel design of the \gls{MSRE} and \gls{MSBR}, the \gls{MSFR} is a pool-type design and
the reactor core consists of a large molten salt pool without graphite moderators to avoid
frequent graphite replacements and positive graphite temperature reactivity feedback. The
\gls{MCFR} is a similar pool-type reactor under active development at TerraPower. TerraPower and
Southern Company embarked on a joint project to design, construct, and operate a prototype
\gls{MCRE} design with funding support from the \gls{DOE}'s \gls{ARDP}. \gls{CAS} launched the
\gls{TMSR} program in 2011 to develop and construct both solid-fueled and liquid-fueled \gls{TMSR}
designs \cite{zou_research_2019}. They finished construction of TMSR-LF1, a 2-MW$_{\text{th}}$
liquid-fueled prototype, in August 2021 and received approval for reactor commissioning in August
2022. Lastly, Canada-based Terrestrial Energy is also developing its \gls{IMSR}, a small modular
\gls{MSR} based on the \gls{MSRE}. It passed a joint technical review by the
Canadian and US nuclear regulators in July 2022.

Developing \gls{MSR} simulation software plays a vital role in supporting \gls{MSR}
development. Accurate reactor modeling capabilities are vital because they
accelerate reactor design and optimization by enabling quicker iteration through numerous design
changes. \gls{MSR} simulation software are also essential tools in reactor safety analysis and
licensing efforts since reactor developers must demonstrate and verify that the \gls{MSR} systems
perform as designed and remain safe under various accident scenarios. 
%
\begin{figure}[htb!]
	\centering
	\includegraphics[width=.7\columnwidth]{msfr}
	\caption{Schematic diagram of the \gls{MSFR}. Retrieved from 
	\cite{allibert_7_2016}.}
	\label{fig:msfr}
\end{figure}

@@

While modeling \glspl{MSR} is not necessarily more difficult than modeling
solid-fueled reactors, we must adapt our software tools to accurately model the
unique phenomena found in these circulating-fuel reactors. The differences in
the challenges of simulating \glspl{MSR} compared to solid-fueled reactors stem
mainly from the liquid fuel form of the fuel salt \cite{diamond_phenomena_2018,
huff_identifying_2019}.

@@

While some \gls{MSR} concepts completely eliminated control rods from their signs, many still
retain them. Control rods consist of strong
neutron absorbers like boron, cadmium, or gadolinium to absorb and thus reduce the number of
neutrons available to contribute to further fission chain reactions. All reactors start operation
with excess reactivity to ensure enough fissile material is available to maintain power operation
and last through to the next scheduled refueling. Control rods and burnable poisons cancel out the
excess reactivity. Control rods are also helpful
for ramping the power up or down during start-up, shut-down, or load-following operations. Most
importantly, control rods provide a safety mechanism for quickly shutting down a reactor under
dangerous conditions. Rapid control rod insertion is the primary mechanism of a reactor scram to
reliably introduce a large negative reactivity insertion in response to an unintended reactivity
insertion or other events that may threaten the reactor's safe operation.

Amongst other tasks in reactor modeling, reactor physics analysts aim to accurately reproduce
control rod worth, which is measured as the relative reduction
in the neutron multiplication factor $k_{eff}$ due to the insertion of the control rod in the
reactor. They are also interested in the accompanying change in flux shape near the
control rod. Neutrons entering the control rods are more likely to be absorbed than
adjacent regions in the reactor core. Therefore, control rods induce highly anisotropic neutron
angular fluxes and sharp gradients in the neutron flux in their vicinity. Control rods also cause
shifts in the neutron energy spectrum because their absorption cross sections are much higher in
the lower neutron energy range; more energetic neutrons generally have a higher probability of
escaping the control rod region. As a consequence, modeling control rods accurately requires
high-fidelity computational methods to capture the angular dependence in the neutron
flux near the highly absorbing medium. Existing reactor analysis workflows often rely on Monte
Carlo methods for high-fidelity neutron transport calculations.

@@

\section{Research Objectives}

The overarching goal of this work is to improve on Moltres as a reliable, intermediate-fidelity
simulation tool for multiphysics \gls{MSR} analysis that can be run on workstations as well as on
larger computing clusters. Thus, new and existing capabilities in Moltres must
be accurate within accepted bounds, rigorously verified, computationally efficient, and highly
scalable. These conditions form the underlying development principles of this work.

@@

\section{Outline}

>>>>

The \gls{MSR} is one of six advanced reactor concepts selected for improved safety, sustainability,
efficiency, and cost over the current generation of predominantly \glspl{LWR} at the \gls{GIF}
\cite{u.s._doe_nuclear_energy_research_advisory_committee_technology_2002}.
Due to the large fuel expansion coefficient, \glspl{MSR} possess an inherently robust
safety feature in the strong negative fuel temperature coefficient of
reactivity \cite{elsheikh_safety_2013}. This reactivity coefficient limits the
maximum temperature that the reactor core would experience in particular accident
scenarios, such as an unprotected reactivity insertion, because the subsequent
rise in core temperatures induces a significant drop in reactivity,
quickly neutralizing the initial reactivity insertion. \glspl{MSR} also
operate at a large thermal margin to boiling and can rely on natural
circulation in the event of a pump failure. For an additional layer of safety, many \gls{MSR}
designs incorporate a drain plug made of actively-cooled frozen salt, which
melts when the core temperatures exceed safety thresholds. Hot molten salt
in the core would then flow into a drain tank designed to hold the fuel salt in
a subcritical configuration to disrupt any further chain fission reactions.

@@

\gls{ORNL} researchers first conceived the \gls{MSR} concept in pursuit of a high-temperature
liquid fuel reactor for the US Aircraft Nuclear Propulsion program in
the 1950s \cite{rosenthal_molten-salt_1970}. They
built the first ever operational \gls{MSR}, the 2.5 MW$_{\text{th}}$
\gls{ARE} reactor at \gls{ORNL}. The successful demonstration of the \gls{ARE} spurred further
research into adapting \glspl{MSR} for civilian power generation \cite{rosenthal_molten-salt_1970}.
Continued \gls{MSR} research efforts culminated in designing constructing, and successfully
operating of the 8-MW$_{\text{th}}$, thermal-spectrum \gls{MSRE} with
graphite channels and a LiF-BeF$_2$-ZrF$_4$-UF$_4$ fuel salt mixture
\cite{haubenreich_experience_1970}. In addition to other operational achievements, the
\gls{MSRE} became the first reactor to run on $^{233}$U fuel bred from $^{232}$Th. Building on
their experience with the \gls{MSRE}, \gls{ORNL} proposed a new program for constructing and
operating a demonstration reactor based on the \gls{MSBR} concept
\cite{macpherson_molten_1985}. The \gls{MSBR} is a thermal-spectrum \gls{MSR} with fertile
$^{232}$Th isotopes mixed directly into the \gls{FLiBe} molten salt for $^{233}$U breeding. The
\gls{MSBR} was to rely on continuous online reprocessing to add fertile
material and remove fission product neutron poisons.
However, the \gls{MSBR} project was canceled prior to the demonstration stage in
favor of the \gls{LMFBR}, which had benefited from a longer development time and more substantial
political backing \cite{macpherson_molten_1985}.

Following a relative lull lasting until the late 1990s, renewed research efforts and interest from
the \gls{GIF}
provided new impetus for \gls{MSR} research and development. As of the end of 2024, numerous
\gls{MSR} designs exist at various stages of development. Leading \gls{MSR} designs, in terms of
development, licensing, or demonstration, include the \gls{MSFR} \cite{merle_optimized_2007},
\gls{MCFR} \cite{terrapower_terrapower_2021}, TMSR-LF1 \cite{zhang_review_2018}, and \gls{IMSR}
\cite{leblanc_18_2017}. The \gls{MSFR} is a fast-spectrum breeder reactor developed through
collaboration among European institutes with funding support from the
European Union. Figure \ref{fig:msfr} shows a schematic diagram of the \gls{MSFR}. As opposed to
the multi-channel design of the \gls{MSRE} and \gls{MSBR}, the \gls{MSFR} is a pool-type design and
the reactor core consists of a large molten salt pool without graphite moderators to avoid
frequent graphite replacements and positive graphite temperature reactivity feedback. The
\gls{MCFR} is a similar pool-type reactor under active development at TerraPower. TerraPower and
Southern Company embarked on a joint project to design, construct, and operate a prototype
\gls{MCRE} design with funding support from the \gls{DOE}'s \gls{ARDP}. \gls{CAS} launched the
\gls{TMSR} program in 2011 to develop and construct both solid-fueled and liquid-fueled \gls{TMSR}
designs \cite{zou_research_2019}. They finished construction of TMSR-LF1, a 2-MW$_{\text{th}}$
liquid-fueled prototype, in August 2021 and received approval for reactor commissioning in August
2022. Lastly, Canada-based Terrestrial Energy is also developing its \gls{IMSR}, a small modular
\gls{MSR} based on the \gls{MSRE}. It passed a joint technical review by
Canadian and US nuclear regulators in July 2022.

Developing \gls{MSR} simulation software plays a vital role in supporting \gls{MSR}
development. Accurate reactor modeling capabilities
accelerate reactor design and optimization by enabling quicker iteration through numerous design
changes. \gls{MSR} simulation software are also essential tools in reactor safety analysis and
licensing efforts since reactor developers must demonstrate and verify that the \gls{MSR} systems
perform as designed and remain safe under various accident scenarios. 
%
\begin{figure}[htb!]
	\centering
	\includegraphics[width=.7\columnwidth]{msfr}
	\caption{Schematic diagram of the \gls{MSFR}. Retrieved from 
	\cite{allibert_7_2016}.}
	\label{fig:msfr}
\end{figure}

@@

While modeling \glspl{MSR} is not necessarily more difficult than modeling
solid-fueled reactors, we must adapt our software tools to accurately model the
unique phenomena found in these circulating-fuel reactors. The differences in
the challenges of simulating \glspl{MSR} compared to solid-fueled reactors stem
primarily from the liquid fuel form of the fuel salt \cite{diamond_phenomena_2018,
huff_identifying_2019}.

@@

While some \gls{MSR} concepts completely eliminated control rods from their signs, many still
retain them. Control rods consist of strong
neutron absorbers like boron, cadmium, or gadolinium to absorb and thus reduce the number of
neutrons available to contribute to further fission chain reactions. All reactors start operation
with excess reactivity to ensure enough fissile material is available to maintain power operation
and last through to the next scheduled refueling. Control rods and burnable poisons balance the
initial excess reactivity. Control rods are also helpful
for ramping the power up or down during start-up, shut-down, or load-following operations. Most
importantly, control rods provide a safety mechanism for quickly shutting down a reactor under
dangerous conditions. Rapid control rod insertion is the primary mechanism of a reactor scram to
reliably introduce a large negative reactivity insertion in response to an unintended reactivity
insertion or other events that may threaten the reactor's safe operation.

Amongst other tasks in reactor modeling, reactor physics analysts aim to accurately reproduce
control rod worth, which is measured as the relative reduction
in the neutron multiplication factor $k_\text{eff}$ due to the insertion of the control rod in the
reactor. They are also interested in the accompanying change in flux shape near the
control rod. Neutrons entering the control rods are more likely to be absorbed than
adjacent regions in the reactor core. Therefore, control rods induce highly anisotropic neutron
angular fluxes and sharp gradients in the neutron flux in their vicinity. Control rods also cause
shifts in the neutron energy spectrum because their absorption cross sections are much higher in
the lower neutron energy range; more energetic neutrons generally have a higher probability of
escaping the control rod region. As a consequence, modeling control rods accurately requires
high-fidelity computational methods to capture the angular dependence in the neutron
flux near the highly absorbing medium. Existing reactor analysis workflows often rely on Monte
Carlo methods for high-fidelity neutron transport calculations.

@@

\section{Research Objectives and Outline}

The overarching goal of this work is to improve on Moltres as a reliable, intermediate-fidelity
simulation tool for multiphysics \gls{MSR} analysis that can be run on workstations as well as on
large leadership-class computing clusters. Thus, new and existing capabilities in Moltres must
be accurate within accepted bounds, rigorously verified, computationally efficient, and highly
scalable. These conditions form the underlying development principles of this work.

@@
